\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-29 chapters/abstract.tex
Second layout run

\chapter*{Abstract}

Individuals with severe motor impairments, such as those with \ac{lis}
or eye-motor dysfunction, face substantial challenges when using traditional
gaze-dependent \acp{bci}.
These systems require users to direct their gaze toward specific targets, a
task that becomes unfeasible for individuals with limited or no eye control.
This work addresses this limitation by developing gaze-independent
\ac{bci} methods, focusing on improving covert \ac{vsa}, where users can direct
their attention to a target without corresponding eye movements.
A key contribution of this work is the compensation for ERP latency jitter, a
variability that negatively impacts decoding performance in covert \ac{vsa}.
By handling this jitter, the proposed method enhances communication accuracy
without requiring gaze control, making \acp{bci} more usable for individuals with
motor impairments.

Beyond gaze independence, the work also advances general \ac{erp} decoding
techniques by refining the structure of linear and multilinear estimators.
These methods improve classification accuracy across a range of \ac{bci}
conditions, particularly when training data are limited.
The introduction of structured regularization in both linear and multilinear
models enhances the interpretability of the classifiers and reduces training
time and computational complexity.
This allows for more efficient training and contributes to generalizability
allowing for more reliable systems that are adaptable to various contexts and
user needs.

@@

>>>>

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Individuals with severe motor impairment, such as those with \acl{lis}, face substantial challenges when using traditional,
gaze-dependent \acp{bci}.
These systems require directing the gaze toward specific targets, which
becomes unfeasible for individuals with limited or no eye control.
This work addresses this limitation by developing gaze-independent
\ac{bci} methods, focusing on improving covert \ac{vsa}, where users can direct
their attention to a target without corresponding eye movements.
A key contribution is the compensation for \ac{erp} latency jitter, a
variability that negatively impacts decoding performance in covert \ac{vsa}.
By handling this jitter, the proposed method enhances communication accuracy
without requiring gaze control, making \acp{bci} more usable for individuals with
motor impairments.

Beyond gaze independence, this work also advances general \ac{erp} decoding
techniques by refining the structure of linear and multilinear estimators.
These methods improve accuracy across a range of \ac{bci}
settings, particularly when training data are limited.
The introduction of structured regularization in both linear and multilinear
models enhances interpretability and reduces training
time and computational complexity.
This allows for more efficient training and contributes to generalizability,
allowing for more reliable systems that are adaptable to various contexts and
user needs.

@@

\bigskip

### 2024-11-29 chapters/abstract_fr.tex
Second layout run

Les individus ayant des handicaps moteurs sévères, tels que ceux atteints de
syndrome d'enfermement ou de dysfonctionnement oculomoteur, rencontrent des défis considérables
lors de l'utilisation d'interfaces cerveau-ordinateur (BCI)  traditionnelles.
Ces systèmes nécessitent que les utilisateurs dirigent leur regard vers des
cibles spécifiques, une tâche impraticable pour les personnes ayant un contrôle oculaire limité ou inexistant.
Ce travail vise \`a pallier cette limitation par améliorer des méthodes de BCI
indépendantes du regard et la décodage de l'attention visuo-spatiale (VSA) cachée, où les utilisateurs peuvent diriger leur attention vers une cible sans mouvements oculaires correspondants.
Une contribution clé de ce travail est la compensation de
la variabilité de latence des potentiels évoqués (ERP), laquelle impacte
la performance de décodage de VSA cachée.
En gérant cette variabilit\'e, la méthode proposée améliore la précision de la communication
sans nécessiter de contrôle du regard, rendant les BCI plus utilisables pour les personnes ayant des handicaps moteurs.

>>>>

Les individus ayant des handicaps moteurs sévères, tels que ceux atteints de
syndrome d'enfermement, rencontrent des défis considérables
lors de l'utilisation d'interfaces cerveau-ordinateur (BCI)  traditionnelles.
Ces systèmes nécessitent de orienter le regard, une tâche impraticable pour
les personnes ayant un contrôle oculaire limité ou inexistant.
Ce travail vise \`a pallier cette limitation en améliorant des méthodes de BCI
indépendantes du regard et en optimisant la décodage de l'attention visuo-spatiale (VSA) cachée,
où l' utilisateurs oriente son attention sans mouvements oculaires correspondants.
Une contribution clé de ce travail est la compensation de
la variabilité de latence des potentiels évoqués (ERP), laquelle impacte
la performance de décodage de VSA cachée.
En gérant cette variabilit\'e, la méthode proposée améliore la précision de la communication
sans nécessiter de contrôle du regard, rendant les BCI plus utilisables pour les personnes ayant des handicaps moteurs.

### 2024-11-29 chapters/abstract_nl.tex
Layout chapter gaze independence
Second layout run

\chapter*{Samenvatting}

Personen met ernstige motorische beperkingen, bijvoorbeeld het Locked-in
Syndroom (LIS) of oog-motorische disfunctie, ondervinden aanzienlijke
uitdagingen bij het gebruik van traditionele visuele brein-computer interfaces (BCI's).
Deze systemen vereisen dat gebruikers hun blik richten op specifieke doelen,
wat onuitvoerbaar wordt voor personen met beperkte of geen controle
over oogbewegingen.
Dit werk richt zich op deze beperking door BCI-methoden onafhankelijk van
oogbewegingen te ontwikkelen, met een focus op het verbeteren van de bedekte
visuospati\"ele aandacht (VSA), de  aandacht voor een doel zonder
overeenkomstige oogbewegingen.
Een belangrijke bijdrage van dit werk is het compenseren van variabiliteit in
de timing van event-gerelateerde potentialen (ERP's), wat de decodeerprestaties
van bedekte VSA negatief beïnvloedt.
Door deze variabiliteit aan te pakken, verbetert de voorgestelde methode de
nauwkeurigheid van communicatie onafhankelijkheid van oogcontrole,
waardoor BCI's bruikbaarder worden voor personen met deze motorische
beperkingen.

Naast onafhankelijkheid van oogbewegingen, verbetert dit werk ook algemene
ERP decodering door de structuur van lineaire en multilineaire technieken te verfijnen.
Deze methoden verbeteren de classificatienauwkeurigheid in een breed scala aan
BCI-condities, met name wanneer trainingsgegevens beperkt zijn.
De introductie van gestructureerde regularisatie in zowel lineaire als
multilineaire modellen verhoogt de interpretateerbaarheid van de classificators
en vermindert de trainingstijd en computationele complexiteit.
Dit maakt efficiëntere training mogelijk en draagt bij aan generalisatie, wat
de ontwikkeling van betrouwbaardere systemen toelaat die zich kunnen
aanpassen aan verschillende contexten en gebruikersbehoeften.

De voorgestelde methoden zijn gevalideerd in experimenten met zowel gezonde
individuen als zeven individuen met een ernstige fysieke beperking en verminderde
oog-motorische controle.
Deze experimenten toonden de robuustheid van de nieuwe decodeermethoden aan,
waarbij werd aangetoond dat het systeem bedekte aandacht effectief kon decoderen,
zelfs wanneer directe blik onmogelijk was.

\paragraph{Trefwoorden:}
brein-computer interface,
electroencephalografie,
event-gerelateerde potentialen,
visuospatiale aandacht,
(multi)lineaire decodering,
fysieke beperkingen

>>>>

\chapter*{Samenvatting}
\addcontentsline{toc}{chapter}{Dutch abstract}

Personen met ernstige motorische beperkingen, bijvoorbeeld het
insluitings\-syndroom, ervaren aanzienlijke uitdagingen bij het gebruik van
traditionele visuele brein-computer interfaces (BCI's).
Deze systemen vereisen gerichte oogbewegingen wat onpraktisch is voor personen
met beperkte of zonder oogbewegingscontrole.
Dit werk ontwikkelt BCI-technieken onafhankelijk van de blik, met nadruk
op bedekte visuospati\"ele aandacht (VSA) -- aandacht zonder overeenkomstige oogbewegingen.
Een belangrijke bijdrage is het compenseren van variabiliteit in de
timing van event-gerelateerde potentialen (ERP's), een factor met een negatieve
impact op de decodering van bedekte VSA.
Door dit aan te pakken, verbetert de nauwkeurigheid van communicatie die
onafhankelijk is van oogcontrole.
wat BCI's bruikbaarder maakt voor personen met oog-motorische beperkingen.

Naast onafhankelijkheid van oogbewegingen, verbetert dit werk ook algemene
ERP decodering door de structuur van lineaire en multilineaire technieken te verfijnen.
Deze methoden verhogen de nauwkeurigheid in verschillende BCI-toepassingen,
vooral wanneer kalibratiegegevens beperkt zijn.
Gestructureerde regularisatie van lineaire en
multilineaire modellen verhoogt de interpretateerbaarheid
en verlaagt de trainingstijd en computationele complexiteit.
Dit draagt bij aan efficiëntere training en generalisatie, wat
de ontwikkeling van betrouwbaardere systemen toelaat, aangepast aan
verschillende contexten en gebruikersbehoeften.

De voorgestelde methoden zijn gevalideerd in experimenten met gezonde
deelnemers en zeven individuen met een ernstige fysieke beperking en verminderde
controle over oogbewegingen.
De resultaten wijzen op de robuustheid van de nieuwe decodeermethoden,
waarbij werd aangetoond dat het systeem bedekte VSA effectief kan decoderen,
zelfs zonder directe blik.

\bigskip

\paragraph{Trefwoorden:}
brein-computer interface,
electroencephalografie,
event-related potentials,
visuospati\"ele aandacht,
(multi)lineaire decodering,
fysieke beperkingen

### 2024-11-29 chapters/conclusion.tex
Second layout run

We evaluated our proposed \ac{wcble} algorithm, its non-iterative
counterpart, and state-of-the-art decoders on the CVSA-ERP dataset as well as on a publicly available
dataset~\cite{Aloise2012} that also contains the overt and covert VSA conditions.
We evaluated the BCI decoding performance in a single-trial classification experiment,
as well as in a target selection experiment reflecting BCI operation.
Performance was significantly different between decoders, but this
result was significantly dependent on the \ac{vsa} condition and the dataset.
While \ac{wcble} was slightly outperformed by the state-of-the-art in overt \ac{vsa}
decoding ($94.28<94.74$ and  $85.25<86.60$ \% \ac{rocauc}),
it increased covert \ac{vsa} decoding ($80.84>78.90$ and $73.38>71.11$ \%
\ac{rocauc}).

@@

To further study the gaze-independent performance of these algorithms, transfer
learning between \ac{vsa} conditions was studied to simulate conditions where an
individual with gaze impairment can end up in different \ac{vsa} settings within
a \ac{bci} operation session due to their lack of proper motor control.
When trained and evaluated on overt VSA data, our proposed \ac{wcble} algorithm
results in a small but significant decrease in performance compared to the state-of-the-art
($85.25>86.60$ and $94.74>94.28$ \% \ac{rocauc}), consistent with the within-condition
results.
For all other pairs of training and evaluation VSA conditions, however,
\ac{wcble} was equal to or significantly better with increases exceeding 4 \%
\ac{rocauc}.

@@

Revisiting the hypotheses put forward in
\cref{sec:gaze-independence/objectives}, we conclude that we have proposed
several techniques that did in effect improve visual oddball \ac{erp}-based
\ac{bci} decoding accuracy, and at least one technique that does so specifically
under conditions relevant for gaze-indepent \ac{bci}.
It was not conclusively proven, however, that these results have a meaningful
impact on on-line \ac{bci} assistive technology use by individuals with
\ac{sspgi}.

>>>>

We evaluated our proposed \ac{wcble} algorithm, its non-iterative
counterpart, and state-of-the-art decoders on the CVSA-ERP dataset as well as on a publicly available
dataset~\cite{Aloise2012} that also contains the overt and covert VSA conditions.
We evaluated the BCI decoding performance in a single-trial classification experiment,
as well as in a target selection experiment reflecting BCI operation.
Performance was significantly different between decoders, but this
result was significantly dependent on the \ac{vsa} condition and the dataset.
While \ac{wcble} was slightly outperformed by the state-of-the-art in overt \ac{vsa}
decoding (94.28<94.74 and  85.25<86.60 \% \acs{rocauc}),
it increased covert \ac{vsa} decoding (80.84>78.90 and 73.38>71.11 \%
\acs{rocauc}).

@@

To further study the gaze-independent performance of these algorithms, transfer
learning between \ac{vsa} conditions was studied to simulate conditions where an
individual with gaze impairment can end up in different \ac{vsa} settings within
a \ac{bci} operation session due to their lack of proper motor control.
When trained and evaluated on overt VSA data, our proposed \ac{wcble} algorithm
results in a small but significant decrease in performance compared to the state-of-the-art
(85.25>86.60 and 94.74>94.28 \% \acs{rocauc}), consistent with the within-condition
results.
For all other pairs of training and evaluation VSA conditions, however,
\ac{wcble} was equal to or significantly better with increases exceeding 4 \%
\ac{rocauc}.

@@

Revisiting the hypotheses put forward in
\cref{sec:gaze-independence/objectives}, we conclude that we have proposed
several techniques that did, in effect, improve visual oddball \ac{erp}-based
\ac{bci} decoding accuracy, and at least one technique that does so specifically
under conditions relevant for gaze-indepent \ac{bci}.
It was not conclusively proven, however, that these results have a meaningful
impact on on-line \ac{bci} assistive technology use by individuals with
\ac{sspgi}.

### 2024-11-29 chapters/covert_align.tex
Second layout run

\Cref{fig:covert-align/single-trial-roc-auc-diff} shows a comparison of \ac{rocauc}
for all pairs of \ac{tlda}, \ac{cble},
and \ac{wcble} for single-trial classification to investigate the contributions of
\ac{cble} and \ac{wcble} relative to their first-stage classifier \ac{tlda}.
For this evaluation, epochs were rejected when the peak-to-peak amplitude
exceeded 800 µV, and for the CVSA-ERP dataset, if the user's gaze differed by more than
10 degrees of visual angle from the fixation crosshair.
\begin{figure}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:covert-align/single-trial-roc-auc-diff}%
		\includegraphics[width=\linewidth]{figures/covert_align/figure4a.pdf}
	\end{subfigure}

	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:covert-align/block-eval}%
		\includegraphics[width=\linewidth]{figures/covert_align/figure4b.pdf}
	\end{subfigure}
  \caption[Decoder performance.]{%
		\subref{fig:covert-align/single-trial-roc-auc-diff} Difference in cross-validated
    single-trial classification \acl{rocauc} ($\Delta$ROC-AUC)
    between \ac{cble}, \ac{wcble}, and their first-stage classifier \ac{tlda}.
    95\% confidence intervals were determined using $k=1000$ bootstrapping.
    Our proposed \ac{wcble} decoder outperforms \ac{tlda} and \ac{cble} for covert and split
    ($d=3$)	\ac{vsa}. \ac{cble} scores on par with \ac{tlda}.
		\subref{fig:covert-align/block-eval} Cross-validated target selection accuracy for
    all decoders plotted as a function of the number of test repetitions in
    different VSA conditions. Significance was determined using one-sided
    (\ac{wcble} $>$ other) Wilcoxon signed-rank tests using False Discovery Rate correction
    ($*= p<0.05$, $**=p<0.01$, $***=p<0.001$). \ac{wcble} generally achieves the highest
    covert \ac{vsa} target selection accuracy.}
\end{figure}

@@

\Cref{fig:covert-align/block-eval} shows the cross-validated \ac{bci} selection
accuracy on the BNCI2014-009 and the CVSA-ERP datasets for all investigated
decoders.
Accuracy was determined by, for each block, selecting the character with the
highest (stage-two if applicable) classifier score and comparing it to the cued target.
Significance was calculated using one-sided Wilcoxon signed-rank tests ($p=0.05$)
corrected for FDR over decoders.
For this evaluation, no epochs were rejected to keep the trial-based structure
of \ac{bci} operation intact.
For all datasets and \ac{vsa} conditions, \ac{cble} scores approximately on par with \ac{tlda}.
Yet, \ac{wcble} yields an improved decoding accuracy for covert \ac{vsa} in both
datasets, which is greatest for smaller numbers of repetitions and
decreases as the number of repetitions increases.
This covert \ac{vsa} accuracy increase over \ac{tlda} is significant in the
BNCI2014-009 dataset for 1 and 3 repetitions, and in CVSA-ERP for 1-5 and 10 repetitions.
Furthermore, while we reported a relative decrease in single-trial \ac{rocauc} for
\ac{wcble} in overt \ac{vsa}, this does not seem to result in a consistent decrease in target
selection accuracy.
No significant increase of \ac{wcble} over other methods was found in split \ac{vsa}.
While Riemannian methods are significantly outperformed by \ac{tlda}, \ac{cble}, and \ac{wcble}
in the BNCI2014-009 dataset, they perform approximately on par with \ac{tlda} and
\ac{wcble} in CVSA-ERP.

@@

\Cref{fig:covert-align/covert-cross-eval} and \cref{fig:covert-align/aloise2012-cross-eval} show
the pairwise differences in area under the ROC curve ($\Delta$ROC-AUC)
between the investigated decoders.
In this evaluation, bad epochs were rejected as in \cref{sec:covert-align/results/bloc-acc}.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:covert-align/covert-cross-eval}
		\includegraphics[width=\linewidth]{figures/covert_align/figure5a.pdf}
	\end{subfigure}

	\bigskip
	\bigskip

	\begin{subfigure}[c]{.48\linewidth}
		\caption{}
		\label{fig:covert-align/aloise2012-cross-eval}
		\includegraphics[width=\linewidth]{figures/covert_align/figure5b.pdf}
	\end{subfigure}\hfill%
	\begin{subfigure}[c]{.48\linewidth}
		\caption{}
		\label{fig:jitter}
		\includegraphics[width=\linewidth]{figures/covert_align/figure5c.pdf}
	\end{subfigure}

  \caption[Cross-condition classifier performance and estimated jitter.]{
		\subref{fig:covert-align/covert-cross-eval},\subref{fig:covert-align/aloise2012-cross-eval}
    Difference in cross-validated area under the receiver
		operating characteristic curves between Classifier-based Latency Estimation
    (CBLE) and Woody CBLE (WCBLE) across conditions for the CVSA-ERP and
    BNCI2014-009 datasets, respectively.
    A decoder is each	time trained on a visuospatial attention (VSA) condition
    and tested on all VSA conditions.
    WCBLE yields an improvement in most
		non-overt \ac{vsa} settings, indicating it is more invariant to eye gaze than
    CBLE and \ac{tlda}.
		\subref{fig:jitter} Jitter characterized as the interquartile range (IQR)
		of target epochs for different \ac{vsa} conditions. Overt \ac{vsa} exhibits lower
    jitter than other conditions.
    Significance of differences
    was determined with two-sided Wilcoxon signed-rank tests with False Discovery
    Rate correction on per-subject jitter ($*= p<0.05$, $**=p<0.01$,
    $***=p<0.001$, $****=p<0.0001$).
	}
\end{figure*}
When comparing \ac{cble} and \ac{tlda}, we do not observe large differences in any of the
evaluated settings, similar to the within-subject conditions, with the greatest
\ac{rocauc} difference -2\% for training in split $(d=2)$ \ac{vsa} and overt \ac{vsa}.
On the contrary, when considering the comparisons between \ac{wcble} and \ac{tlda}, we
see that performance is on par or greater using \ac{wcble} for most conditions,
except for within-overt \ac{vsa} decoding, with the greatest \ac{rocauc}
difference +4\% for training in split ($d=2$) \ac{vsa} and testing in overt
\ac{vsa}.

\subsection{Jitter analysis}
\label{sec:covert-align/results/jitter}
Finally, an analysis is performed to quantify the jitter presence in the
different \ac{vsa} conditions.
To obtain comparable results across \ac{vsa} conditions, \ac{wcble} was
evaluated per session and trained on all combined \ac{vsa} conditions as in
\cref{sec:covert-align/results/cross}.
Since all conditions have the P3 in common, the estimated latencies can be
interpreted as P3 latencies.
Results are shown in \cref{fig:jitter}.

>>>>

\Cref{fig:covert-align/single-trial-roc-auc-diff} shows a comparison of \ac{rocauc}
for all pairs of \ac{tlda}, \ac{cble},
and \ac{wcble} for single-trial classification to investigate the contributions of
\ac{cble} and \ac{wcble} relative to their first-stage classifier \ac{tlda}.
For this evaluation, epochs were rejected when the peak-to-peak amplitude
exceeded 800 µV, and for the CVSA-ERP dataset, if the user's gaze differed by more than
10 degrees of visual angle from the fixation crosshair.
\begin{figure}[t]
	%\includegraphics[width=\linewidth]{figures/covert_align/figure4a.pdf}
  \input{figures/covert_align/roc_auc_diff.pgf}
  \caption[Difference in classification \acs{rocauc}.]{%
    Difference in cross-validated
    single-trial classification \acl{rocauc} (\%.)
    between \ac{cble}, \ac{wcble}, and their first-stage classifier \ac{tlda}.
    95\% confidence intervals were determined using $k=1000$ bootstrapping.
    Our proposed \ac{wcble} decoder outperforms \ac{tlda} and \ac{cble} for covert and split
    ($d=3$)	\ac{vsa}. \ac{cble} scores on par with \ac{tlda}.
    }
		\label{fig:covert-align/single-trial-roc-auc-diff}%
\end{figure}

@@

\Cref{fig:covert-align/block-eval} shows the cross-validated \ac{bci} selection
accuracy on the BNCI2014-009 and the CVSA-ERP datasets for all investigated
decoders.
Accuracy was determined by, for each block, selecting the character with the
highest (stage-two if applicable) classifier score and comparing it to the cued target.
Significance was calculated using one-sided Wilcoxon signed-rank tests ($p=0.05$)
corrected for FDR over decoders.
For this evaluation, no epochs were rejected to keep the trial-based structure
of \ac{bci} operation intact.
For all datasets and \ac{vsa} conditions, \ac{cble} scores approximately on par with \ac{tlda}.
Yet, \ac{wcble} yields an improved decoding accuracy for covert \ac{vsa} in both
datasets, which is greatest for smaller numbers of repetitions and
decreases as the number of repetitions increases.
This covert \ac{vsa} accuracy increase over \ac{tlda} is significant in the
BNCI2014-009 dataset for 1 and 3 repetitions, and in CVSA-ERP for 1-5 and 10 repetitions.
Furthermore, while we reported a relative decrease in single-trial \ac{rocauc} for
\ac{wcble} in overt \ac{vsa}, this does not seem to result in a consistent decrease in target
selection accuracy.
No significant increase of \ac{wcble} over other methods was found in split \ac{vsa}.
While Riemannian methods are significantly outperformed by \ac{tlda}, \ac{cble}, and \ac{wcble}
in the BNCI2014-009 dataset, they perform approximately on par with \ac{tlda} and
\ac{wcble} in CVSA-ERP.
\begin{figure}[t]
		\label{fig:covert-align/block-eval}%
		\includegraphics[width=\linewidth]{figures/covert_align/figure4b.pdf}
  \caption[Selection accuracy.]{%
		Cross-validated target selection accuracy for
    all decoders plotted as a function of the number of test repetitions in
    different VSA conditions. Significance was determined using one-sided
    (\ac{wcble} $>$ other) Wilcoxon signed-rank tests using False Discovery Rate correction
    ($*= p<0.05$, $**=p<0.01$, $***=p<0.001$). \ac{wcble} generally achieves the highest
    covert \ac{vsa} target selection accuracy.}
\end{figure}

@@

\Cref{fig:covert-align/covert-cross-eval} and \cref{fig:covert-align/aloise2012-cross-eval} show
the pairwise differences in area under the ROC curve ($\Delta$ROC-AUC)
between the investigated decoders.
In this evaluation, bad epochs were rejected as in \cref{sec:covert-align/results/bloc-acc}.
\begin{figure}[t!]
  \sffamily
  \footnotesize
	\begin{subfigure}{\textwidth}
		%\includegraphics[width=\linewidth]{figures/covert_align/figure5a.pdf}
    \hspace{-.5in}\input{figures/covert_align/cross_cvsa-erp.pgf}
		\caption{Cross-condition performance for CVSA-ERP}
		\label{fig:covert-align/covert-cross-eval}
	\end{subfigure}
	\begin{subfigure}{\textwidth}
		%\includegraphics[width=\linewidth]{figures/covert_align/figure5b.pdf}
    \hspace{-.5in}\input{figures/covert_align/cross_bnci2014-009.pgf}
		\caption{Cross-condition performance for BNCI2014-009}
		\label{fig:covert-align/aloise2012-cross-eval}
	\end{subfigure}%
  \caption[Cross-condition classifier performance]{
    Difference in cross-validated  \ac{rocauc} (\%.) between Classifier-based Latency Estimation
    (CBLE) and Woody CBLE (WCBLE) across conditions for the CVSA-ERP and
    BNCI2014-009 datasets, respectively.
    A decoder is each	time trained on a visuospatial attention (VSA) condition
    and tested on all VSA conditions.
    WCBLE yields an improvement in most
		non-overt \ac{vsa} settings, indicating it is more invariant to eye gaze than
    CBLE and \ac{tlda}.
  }
\end{figure}
When comparing \ac{cble} and \ac{tlda}, we do not observe large differences in any of the
evaluated settings, similar to the within-subject conditions, with the greatest
\ac{rocauc} difference -2\% for training in split $(d=2)$ \ac{vsa} and overt \ac{vsa}.
On the contrary, when considering the comparisons between \ac{wcble} and \ac{tlda}, we
see that performance is on par or greater using \ac{wcble} for most conditions,
except for within-overt \ac{vsa} decoding, with the greatest \ac{rocauc}
difference +4\% for training in split ($d=2$) \ac{vsa} and testing in overt
\ac{vsa}.

\subsection{Jitter analysis}
\label{sec:covert-align/results/jitter}
Finally, an analysis is performed to quantify the jitter presence in the
different \ac{vsa} conditions.
To obtain comparable results across \ac{vsa} conditions, \ac{wcble} was
evaluated per session and trained on all combined \ac{vsa} conditions as in
\cref{sec:covert-align/results/cross}.
Since all conditions have the P3 in common, the estimated latencies can be
interpreted as P3 latencies.
Results are shown in \cref{fig:jitter}.
\begin{figure}[t!]
		%\includegraphics[width=\linewidth]{figures/covert_align/figure5c.pdf}
    \input{figures/covert_align/jitter.pgf}
    \caption[Estimated jitter]{ Jitter characterized as the interquartile range (IQR)
	of target epochs for different \ac{vsa} conditions. Overt \ac{vsa} exhibits lower
   jitter than other conditions.
   Significance of differences
   was determined with two-sided Wilcoxon signed-rank tests with False Discovery
   Rate correction on per-subject jitter ($*= p<0.05$, $**=p<0.01$,
   $***=p<0.001$, $****=p<0.0001$).}
		\label{fig:jitter}
\end{figure}

### 2024-11-29 chapters/patients.tex
Second layout run

To make the interface suitable for use by individuals with
\ac{sspi}~\cite{FriedOken2020}, the
number of blocks was decreased to 6 per \ac{vsa} setting.
\Ac{isi} was increased to $200\pm50$ms to decrease task difficulty.
The experiment also started with a training block in each condition, where the
participant was instructed with feedback on their performance to ensure they
understood and were able to perform the task.

@@

However, \ac{wcble} did not improve \ac{tlda} performance in the free \ac{vsa} setting, but
XDAWNCov+TS+LDA performance was slightly lower here (though not
significantly).
(\ac{wcble} 74.15\%, XDAWNCov+TS+LDA 71.88\%, \ac{tlda} 74.27\%).
More interestingly, we noticed that performances of the decoders in free
\ac{vsa} were close to those in the overt \ac{vsa}.
A substantial decrease in performance from the overt setting to the free
setting was observed for subjects PC3
(\ac{wcble}: $70.31\%>62.14\%$, XDAWNCov+TS+LDA: $65.78\%>62.18\%$, \ac{tlda}:
$70.49\%>63.76\%$)
and PC4
(\ac{wcble}: $65.56\%>55.71\%$, XDAWNCov+TS+LDA: $62.02\%>54.24\%$, \ac{tlda}:
$66.12\%>57.08\%$).
For PB2, who also relied on covert \ac{vsa} during the uncued free \ac{vsa}
according to gaze tracking setting, the decrease in performance was also
present, but not as substantial
(\ac{wcble}: $82.76\%>78.88\%$, XDAWNCov+TS+LDA: $80.74\%>77.99\%$, \ac{tlda}:
$83.21\%>78.84\%$).

>>>>

To make the interface suitable for use by individuals with
\ac{sspi}~\cite{FriedOken2020}, the
number of blocks was decreased to 6 per \ac{vsa} setting.
\Ac{isi} was increased to 200$\pm$50 ms to decrease task difficulty.
The experiment also started with a training block in each condition, where the
participant was instructed with feedback on their performance to ensure they
understood and were able to perform the task.

@@

However, \ac{wcble} did not improve \ac{tlda} performance in the free \ac{vsa} setting, but
XDAWNCov+TS+LDA performance was slightly lower here (though not
significantly).
(\ac{wcble} 74.15\%, XDAWNCov+TS+LDA 71.88\%, \ac{tlda} 74.27\%).
More interestingly, we noticed that performances of the decoders in free
\ac{vsa} were close to those in the overt \ac{vsa}.
A substantial decrease in performance from the overt setting to the free
setting was observed for subjects PC3
(\ac{wcble}: 70.31>62.14 \%, XDAWNCov+TS+LDA: 65.78>62.18 \%, \ac{tlda}:
70.49>63.76 \%)
and PC4
(\ac{wcble}: 65.56>55.71 \%, XDAWNCov+TS+LDA: 62.02>54.24 \%, \ac{tlda}:
66.12>57.08 \%).
For PB2, who also relied on covert \ac{vsa} during the uncued free \ac{vsa}
according to gaze tracking setting, the decrease in performance was also
present, but not as substantial
(\ac{wcble}: 82.76>78.88 \%, XDAWNCov+TS+LDA: 80.74>77.99 \%, \ac{tlda}:
83.21>78.84 \%).

### 2024-11-29 chapters/patients_table.tex
Second layout run

\sffamily
\small
\begin{tabular}{@{}llrrrlllrr@{}}
  \toprule
  \textbf{ID}  & \textbf{Diagnosis} & \textbf{Age} & \textbf{Sex} & \textbf{Hand.} &
  \textbf{Speech} & \textbf{Trach.} & \textbf{Communication} &
  \textbf{W} & \textbf{KB} \\ \midrule
  PA1 & bulbar-onset \acs{als} & 58  & M   & L     & anarthric  & no          & tablet                 & 3  & 4 \\
  PB1 & \acs{fa} & 41  & M   & L     & dysarthric & no          & verbal                 & 3  & 3 \\
  PB2 & \acs{fa} & 43  & F   & R     & dysarthric & no          & verbal                 & 3  & 3 \\
  PB4 & \acs{fa} & 48  & M   & R     & dysarthric & no          & verbal                 & 3  & 3 \\
  PC2 & brainstem stroke & 43  & M   & R     & anarthric  & yes         &  \makecell[l]{prompting\\+eye movement} & 2 & 4  \\
  PC3 & brainstem stroke & 43  & F   & R     & anarthric  & yes         & letterboard            & 2  & 3 \\
  PC4 & \makecell[l]{left cerebellar stroke \\ (trombosis of the  basilar \\ artery)} & 54  & M   & R     & anarthric  & yes & letterboard & 2 & 3 \\
  \bottomrule
\end{tabular}

>>>>

\sffamily
\footnotesize
\begin{tabular}{@{}llrrrlllrr@{}}
  \toprule
  \textbf{ID}  & \textbf{Diagnosis} & \textbf{Age} & \textbf{Sex} & \textbf{Hand.} &
  \textbf{Speech} & \textbf{Trach.} & \textbf{Communication} &
  \textbf{W} & \textbf{KB} \\ \midrule
  PA1 & bulbar-onset \acs{als} & 58  & M   & L     & anarthric  & no          & tablet                 & 3  & 4 \\
  PB1 & \acs{fa} & 41  & M   & L     & dysarthric & no          & verbal                 & 3  & 3 \\
  PB2 & \acs{fa} & 43  & F   & R     & dysarthric & no          & verbal                 & 3  & 3 \\
  PB4 & \acs{fa} & 48  & M   & R     & dysarthric & no          & verbal                 & 3  & 3 \\
  PC2 & brainstem stroke & 43  & M   & R     & anarthric  & yes         &  \makecell[l]{prompting\\+eye movement} & 2 & 4  \\
  PC3 & brainstem stroke & 43  & F   & R     & anarthric  & yes         & letterboard            & 2  & 3 \\
  PC4 & \makecell[l]{left cerebellar stroke \\ (trombosis of the  \\ basilar  artery)} & 54  & M   & R     & anarthric  & yes & letterboard & 2 & 3 \\
  \bottomrule
\end{tabular}
